---
title: "Fair and Private Clinical Modeling: A Two-Package Workflow"
author: "Cuiwei Gao"
date: "2026-04-19"
categories: [R, Clinical AI, Fairness, Privacy, CRAN]
description: "A practical workflow combining syntheticdata and clinicalfair to generate privacy-preserving clinical datasets and audit prediction models for group fairness."
---

Clinical prediction models are increasingly used to triage patients, flag
deterioration risk, and allocate scarce resources. Two problems shadow
every deployment: **privacy** (can the training data be shared safely
across institutions?) and **fairness** (does the model perform equitably
across demographic groups?). These concerns are usually tackled separately,
but they interact — you cannot audit a model for fairness if privacy rules
prevent you from accessing the data it was trained on.

This post walks through a combined workflow using two R packages:
[syntheticdata](https://cran.r-project.org/package=syntheticdata) for
generating and validating privacy-preserving synthetic clinical data, and
[clinicalfair](https://cran.r-project.org/package=clinicalfair) for
auditing the resulting prediction model for group fairness. Every function
call below uses real exported functions from these packages.

## Step 1: Generate Synthetic Data

Suppose you have a clinical dataset with patient demographics, lab values,
and a binary outcome (e.g., 30-day readmission). You want to share a
synthetic copy with an external collaborator. `synthesize()` supports
three methods: Gaussian copula, bootstrap resampling, and Laplace noise
injection.

```{r}
#| eval: false
library(syntheticdata)

# Original clinical dataset (not shared externally)
data_original <- read.csv("clinical_cohort.csv")

# Generate synthetic data using Gaussian copula
data_synthetic <- synthesize(
  data = data_original,
  method = "copula",
  n = nrow(data_original)
)
```

The copula method estimates the joint distribution of all variables via
their empirical marginals and a Gaussian copula, then samples from it.
This preserves correlation structure while breaking the one-to-one link
between synthetic and real records.

## Step 2: Validate the Synthetic Data

Before trusting synthetic data for downstream analysis, you need to verify
that it preserves the statistical properties of the original.
`validate_synthetic()` runs a suite of distributional comparisons:

```{r}
#| eval: false
validation <- validate_synthetic(
  original = data_original,
  synthetic = data_synthetic
)
validation
```

The output reports per-variable KS statistics for continuous columns and
chi-squared statistics for categorical columns, along with an overall
utility score. If a variable diverges substantially, you know to
investigate before proceeding.

## Step 3: Assess Privacy Risk

Even well-constructed synthetic data can leak information about
individuals in the original dataset, especially for rare combinations of
attributes. `privacy_risk()` performs membership inference testing and
attribute disclosure risk assessment:

```{r}
#| eval: false
risk <- privacy_risk(
  original = data_original,
  synthetic = data_synthetic
)
risk
```

The membership inference test trains a classifier to distinguish real
from synthetic records; if it cannot do much better than chance, the
synthetic data does not obviously memorize the training set. The attribute
disclosure component checks whether knowing a subset of quasi-identifiers
in the synthetic data allows reconstructing sensitive attributes from the
original. Together, these provide a practical privacy profile you can
report to your IRB or data governance board.

## Step 4: Train a Prediction Model

With a validated, privacy-assessed synthetic dataset in hand, you (or your
collaborator) can train a clinical prediction model. For this example we
use a simple logistic regression, but the downstream fairness audit is
model-agnostic.

```{r}
#| eval: false
# Train on synthetic data
model <- glm(
  readmission ~ age + sex + race + lab_value_1 + lab_value_2 + comorbidity_index,
  data = data_synthetic,
  family = binomial
)

# Generate predictions
data_synthetic$pred_prob <- predict(model, type = "response")
data_synthetic$pred_class <- ifelse(data_synthetic$pred_prob > 0.5, 1, 0)
```

## Step 5: Prepare Data for Fairness Audit

`fairness_data()` bundles the observed outcomes, predicted scores, and
protected attributes into a structure that all downstream clinicalfair
functions expect:

```{r}
#| eval: false
library(clinicalfair)

fdata <- fairness_data(
  data = data_synthetic,
  outcome = "readmission",
  predicted = "pred_prob",
  group = "race"
)
```

## Step 6: Compute Group-Stratified Metrics

`fairness_metrics()` computes performance metrics stratified by the
protected attribute. By default it calculates sensitivity, specificity,
PPV, NPV, AUC, and calibration slope for each group, with bootstrap
confidence intervals:

```{r}
#| eval: false
metrics <- fairness_metrics(fdata)
metrics
```

The output is a tidy data frame — one row per group per metric — that
you can pipe directly into ggplot2 for visualization or into reporting
templates.

## Step 7: Four-Fifths Rule Check

The four-fifths (or 80%) rule, borrowed from employment discrimination
law and increasingly applied to clinical AI, flags a metric as disparate
if the worst-performing group's rate falls below 80% of the
best-performing group's rate. `fairness_report()` automates this check:

```{r}
#| eval: false
report <- fairness_report(fdata)
report
```

The report identifies which metrics violate the four-fifths rule, which
group comparisons drive the violation, and the magnitude of the disparity.
This gives you a concrete, auditable summary to include in model
documentation or regulatory submissions.

## Step 8: Intersectional Analysis

Single-axis fairness (stratifying by race alone, or sex alone) can mask
compounded disparities at the intersection of multiple attributes.
`intersectional_fairness()` crosses protected groups and repeats the
analysis:

```{r}
#| eval: false
inter <- intersectional_fairness(
  data = data_synthetic,
  outcome = "readmission",
  predicted = "pred_prob",
  groups = c("race", "sex")
)
inter
```

This might reveal, for example, that the model performs adequately for
each race group and each sex group in isolation, but fails for a specific
race–sex intersection. These are the disparities that single-axis audits
miss.

## Step 9: Threshold-Based Mitigation

If disparities are identified, one practical mitigation strategy is to
use group-specific decision thresholds that equalize a chosen metric
(e.g., sensitivity) across groups. `threshold_optimize()` searches for
the threshold set that achieves this:

```{r}
#| eval: false
optimized <- threshold_optimize(
  fdata,
  metric = "sensitivity"
)
optimized
```

The output includes per-group thresholds and the resulting equalized
metric values. This is not a silver bullet — adjusting thresholds trades
off one metric against another — but it provides a transparent, auditable
mechanism for reducing disparity, which is often what regulatory reviewers
want to see.

## Putting It Together

The combined workflow — generate synthetic data, validate it, assess
privacy risk, train a model, and audit for fairness — addresses the two
core concerns in a single reproducible pipeline. The synthetic data lets
you share and collaborate without exposing patient records; the fairness
audit ensures the resulting model does not quietly disadvantage
vulnerable subgroups.

Both packages are on CRAN:

```{r}
#| eval: false
install.packages("syntheticdata")
install.packages("clinicalfair")
```

- **syntheticdata:** [CRAN](https://cran.r-project.org/package=syntheticdata) |
  [GitHub](https://github.com/CuiweiG/syntheticdata) |
  [Docs](https://cuiweig.github.io/syntheticdata)
- **clinicalfair:** [CRAN](https://cran.r-project.org/package=clinicalfair) |
  [GitHub](https://github.com/CuiweiG/clinicalfair) |
  [Docs](https://cuiweig.github.io/clinicalfair)
