---
title: "Introducing lineagefreq: Tracking Pathogen Variant Dynamics in R"
author: "Cuiwei Gao"
date: "2026-04-16"
categories: [R, Bioinformatics, Genomic Surveillance, CRAN]
description: "A walkthrough of lineagefreq, an R package for modelling pathogen lineage frequencies, estimating growth advantages, and forecasting variant trajectories from genomic surveillance data."
---

When a new pathogen variant begins circulating, one of the first questions
public health agencies ask is: *how fast is it growing relative to existing
lineages?* Answering that question from raw sequence counts is harder than
it looks. Counts are noisy, sampling is uneven across time and geography,
and the multinomial nature of the data — every sequence belongs to exactly
one lineage — means you cannot just fit separate logistic curves and call
it a day.

[lineagefreq](https://cran.r-project.org/package=lineagefreq) is an R
package that tackles this problem end to end. It fits multinomial logistic
regression models to genomic surveillance count data, estimates relative
growth advantages between lineages, generates short-term frequency
forecasts, and validates those forecasts with rigorous rolling-origin
backtesting. The current CRAN release is 0.2.0; the development version
(0.6.0) on [GitHub](https://github.com/CuiweiG/lineagefreq) adds several
new features.

## Five Engines, One Interface

A key design decision in lineagefreq is offering multiple estimation
engines behind a single `fit_model()` interface. You choose the engine
with one argument; everything else — data format, output structure,
downstream methods — stays the same.

The five engines are:

- **`mlr`** — standard multinomial logistic regression (frequentist,
  fast, good default)
- **`hier_mlr`** — hierarchical multinomial logistic regression for
  multi-site or multi-region data
- **`piantham`** — the Piantham et al. method, commonly used in
  SARS-CoV-2 variant tracking
- **`fga`** — fixed growth advantage model, assumes constant relative
  fitness over the estimation window
- **`garw`** — growth advantage random walk, allows the fitness advantage
  to drift over time (Bayesian)

This means you can benchmark multiple statistical approaches on the same
dataset with minimal code changes, which is exactly what you want when
advising decision-makers who need to understand model uncertainty.

## Built-in Real-World Datasets

Rather than shipping toy data, lineagefreq includes four real CDC and
public surveillance datasets:

- `cdc_sarscov2_jn1` — SARS-CoV-2 JN.1 sublineage emergence data
- `cdc_ba2_transition` — the BA.1-to-BA.2 transition in the United States
- `influenza_h3n2` — influenza H3N2 clade competition data
- `sarscov2_us_2022` — US Omicron sublineage dynamics across 2022

These serve double duty: they make vignettes and examples reproducible,
and they provide realistic test cases for validating new methods. The
BA.2 transition dataset, for example, is the basis for the validation
result I discuss below.

## A Quick Example

Here is a minimal workflow using the BA.2 transition data. We fit a model,
extract growth advantages, and generate a short-term forecast:

```{r}
#| eval: false
library(lineagefreq)

# Load the BA.1 → BA.2 transition dataset
data("cdc_ba2_transition")

# Fit a multinomial logistic regression model
fit <- fit_model(
  data = cdc_ba2_transition,
  engine = "mlr"
)

# Estimate growth advantages relative to the reference lineage
ga <- growth_advantage(fit)
ga
```

The `growth_advantage()` output includes point estimates and confidence
intervals. For the BA.2 versus BA.1 comparison, lineagefreq estimates a
growth advantage of approximately 1.34×, which aligns well with published
estimates in the literature (typically reported in the range of 1.3–1.5×).

Generating a forecast and evaluating it is equally straightforward:

```{r}
#| eval: false
# 4-week-ahead frequency forecast
fc <- forecast(fit, horizon = 4)

# Rolling-origin backtesting with proper scoring
bt <- backtest(
  data = cdc_ba2_transition,
  engine = "mlr",
  horizon = 4,
  window = 8
)

# Score the backtested forecasts
scores <- score_forecasts(bt)
scores
```

The `backtest()` function walks a sliding window across the time series,
refitting the model at each origin and scoring out-of-sample predictions.
`score_forecasts()` computes proper scoring rules (log score, Brier score)
so you can compare engines or parameter choices on a level playing field.

## Tidy Integration

lineagefreq fits naturally into the tidyverse ecosystem. Fitted model
objects support the standard [broom](https://broom.tidymodels.org/)
generics:

```{r}
#| eval: false
library(broom)

# Coefficient-level summaries
tidy(fit)

# Model-level goodness-of-fit statistics
glance(fit)

# Observation-level fitted values and residuals
augment(fit)
```

This makes it easy to pipe results into ggplot2 for visualization or into
dplyr workflows for further analysis, without learning a new set of
accessor functions.

## Eight Vignettes

The package ships with eight vignettes covering everything from a
getting-started guide to advanced topics like hierarchical multi-region
modelling and custom scoring rule implementation. The vignettes are built
around the real datasets and walk through complete analysis pipelines, not
just isolated function calls. You can browse them on the
[pkgdown site](https://cuiweig.github.io/lineagefreq).

## What Comes Next

The development version (0.6.0) on GitHub is where active work happens.
Current priorities include improving the Bayesian engine performance,
adding support for weighted observation models, and expanding the forecast
evaluation toolkit. If you work in genomic surveillance or infectious
disease modelling and want to try it out, I would love to hear your
feedback.

- **CRAN:** <https://cran.r-project.org/package=lineagefreq>
- **GitHub:** <https://github.com/CuiweiG/lineagefreq>
- **pkgdown:** <https://cuiweig.github.io/lineagefreq>
- **Posit Community tutorial:** [From Sequence Counts to Variant Forecasts](https://forum.posit.co/t/from-sequence-counts-to-variant-forecasts-a-lineagefreq-tutorial/211035)

```{r}
#| eval: false
install.packages("lineagefreq")

# Or the development version:
# pak::pak("CuiweiG/lineagefreq")
```
